---
title: "W203 Statistics - Lab 1"
author: "Justine Heritage, Morris Burkhardt, Daniel Volk"
date: "May 31, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This analysis is motivated by the following research question:

> What is the relationship between CEO salary and company performance?

Our data was provided with the following codebook:

Variable Name | Variable Meaning
------------- | ----------------
salary        | 1990 compensation, $1000s
age           | in years
college       | =1 if attended college
grad          | =1 if attended graduate school
comten        | years with company
ceoten        | years as CEO with company
profits       | 1990 profits, millions
mktval        | market value, end 1990, millions

As we look at these variables, we need to define how we will measure company performance and consider possible limitations in this dataset.

Our primary indicators of company performance will be profit and market value. However, there are several other factors than CEO salary that could influence a company's performance. For example:

* Performance of the CEO's predecessor
* Market environment of a company's industry or sector
* Performance of executive team and employees
* Influence from board of directors
* Prior year profits

To address some of these issues, we will also examine the length of the CEO's tenure. However, a lack of knowledge about external factors will prevent us from making any claims of causality.

# Setup

To begin our analysis, we used the `car` library and loaded the provided data set.

```{r setup_car}
library(car)
load("ceo_w203.RData")
```

We take an initial look at our data set.

```{r initial}
head(CEO)
names(CEO)
str(CEO)
```

There are 185 observations over 8 variables. We notice that `college` and `grad` are dummy variables. The rest of the variables are numeric. The `salary` variable is measured in millions of \$, `profits` and `mktval` are measured in thousands of \$. The `age`, `comten` (years the CEO has been with the company) and `ceoten` (years as CEO with the company) variables are integer values.

## Data Selection

We summarize the data.

```{r summary1}
summary(CEO)
```

We notice that there is an unusual minimum, $-1$, for `mktval`. We determine that in both `mktval` and `profits`, these observations have missing data. We replace the $-1$s with `NA`.

```{r replace_neg1}
CEO$mktval[CEO$mktval==-1] <- NA
CEO$profits[CEO$profits==-1] <- NA
```

By looking at the tail of our data, we also notice that row 184 and 185 might be duplicates. All values - besides age - are identical. These two rows are also the only rows that indicate that the CEO went to grad school but not college. It's possible that this data is inconsistent.

```{r duplicates}
# final 2 values look duplicated

tail(CEO[order(as.numeric(row.names(CEO))),],2)

# potentially synthesize rows into 1 with age value 31.5
```


# Exploratory Analysis

## Univariate Analysis

First, we look at the histogram for `salary`.

```{r salary_hist}
hist(CEO$salary, breaks = 20, main = "Histogram of CEO salaries, in $1000s" , 
     xlab = NULL)
```


We notice that the data is positively skewed and there are some large outliers.

Next, we look at the histogram for `profits`.


```{r profits_hist}
hist(CEO$profits, breaks = 30, main = "Histogram of Company Profits, in million $" , 
     xlab = NULL)
```

The `profits` variable has both negative values and outliers at the high end.

Finally, we look at the histogram form `mktval`.

```{r mktval_hist}
hist(CEO$mktval, breaks = 20, main = "Histogram of Company Market Value, in million $" , 
     xlab = NULL)
```

Again, we see large outliers in the `mktval` variable.

Since all distributions are heavily skewed with outliers to the far right, we perform a logarithmic transformation. For the `profits` variable, we also have to consider omitting all values $<= 0$ in the transformation.

First we check how many profits values actually are zero or negative.

```{r}
sum(CEO$profits<=0, na.rm = TRUE)
```

We would be omitting 10 values out of 185, which is less than 6% of our data.

Next, we look at histograms for our log-transformed data.

```{r}
CEO$log_salary  = log10(CEO$salary)
CEO$log_profits = log10(CEO$profits)
CEO$log_mktval  = log10(CEO$mktval)

hist(CEO$log_salary, breaks = 20)
hist(CEO$log_profits, breaks = 20)
hist(CEO$log_mktval, breaks = 20)
```

The negative profit values are transformed into NAs. 

The transformed `salary` and `profits` variables resemble the normal distribution. The `mktval` variable still has a minor left skew, but with no significant outliers.


# Bivariate analysis

## Salary, Market Value and Profits

First, we look at the linear correlations between `salary` and `profits` and `salary` and `mktval`. To better compare the two correlation coefficients, we'll only look at positive profit values. Then, we look at a scatterplot matrix for these three variables.

```{r}
pos_profits = CEO$profits > 0

cor(CEO$profits[pos_profits], CEO$salary[pos_profits], use = "complete.obs")
cor(CEO$mktval, CEO$salary, use = "complete.obs")
cor(CEO$profits[pos_profits], CEO$mktval[pos_profits], use = "complete.obs")
```

```{r}
scatterplotMatrix( ~ salary + profits + mktval, data = CEO, 
 #                reg.line="" , 
                   smoother="",
                   diagonal = "histogram", 
                   main = "Scatterplot Matrix for original data")
```

There is a moderate positive linear correlation between both `salary` and both `profits` and `mktval`. We also notice a strong linear correlatio between `profits` and `mktval`.

Next we'll look at the correlations between the log-transformed variables and their scatterplot matrix. 

```{r}
cor(CEO$log_profits, CEO$log_salary, use = "complete.obs")
cor(CEO$log_mktval, CEO$log_salary, use = "complete.obs")
cor(CEO$log_profits, CEO$log_mktval, use = "complete.obs")
```

```{r}
scatterplotMatrix( ~ log_salary + log_profits + log_mktval, data = CEO, 
 #                reg.line="" , 
                   smoother="",
                   diagonal = "histogram", 
                   main = "Scatterplot Matrix for transformed data")
```

We can see a stronger relationship between `salary` and `profits` as well as `salary` and `mktval` with the correlation increasing from `r cor(CEO$profits[pos_profits], CEO$salary[pos_profits], use = "complete.obs")` to `r cor(CEO$log_profits, CEO$log_salary, use = "complete.obs")` and `r cor(CEO$mktval, CEO$salary, use = "complete.obs")` to `r cor(CEO$log_mktval, CEO$log_salary, use = "complete.obs")`, respectively. Due to the nonlinearity of the relationship, the calculated correlation on the original (non-transformed) variables underestimates the actual relationship.

<!-- Let's now take a look at the scatterplot matrix for the key variables under examination. Note that all data points, where profits is negative, are omitted. -->


<!-- ```{r} -->
<!-- scatterplotMatrix( ~ salary[profits <= 0] + profits[profits <= 0] + mktval[profits <= 0],  -->
<!--                    data = CEO,  -->
<!--  #                reg.line="" ,  -->
<!--                    smoother="", -->
<!--                    diagonal = "histogram",  -->
<!--                    main = "Scatterplot Matrix for negative profit values") -->
<!-- ``` -->

# Further considerations

## CEO Tenure

To limit the effect of the performance of past CEOs on company performance, we consider weighting `profits` and `mktval` based on CEO tenure (`ceoten`) and calculating the correlations.

```{r}
CEO$weighted_profits = CEO$profits * CEO$ceoten
CEO$weighted_mktval = CEO$mktval * CEO$ceoten

cor(CEO$salary, CEO$weighted_profits, use = "complete.obs")
cor(CEO$salary, CEO$weighted_mktval, use = "complete.obs")
cor(CEO$weighted_profits, CEO$weighted_mktval, use = "complete.obs")
```


Since there are `ceoten` values that are zero, some of the weighted values are now zero. We replace those with NAs before logarithmic transformation. We can justify omitting those values by assuming that CEOs who have been leading a company for less than a year do not yet have a significant effect on the company's performance.  

We perform log transforms on our weighted variables, recalculate correlations, and draw histograms for both.


```{r}
CEO$log_weighted_profits = log10(CEO$weighted_profits)
CEO$log_weighted_mktval  = log10(CEO$weighted_mktval)

CEO$log_weighted_profits[CEO$weighted_profits == 0] = NA
CEO$log_weighted_mktval[CEO$weighted_mktval == 0] = NA

cor(CEO$salary, CEO$log_weighted_profits, use = "complete.obs")
cor(CEO$salary, CEO$log_weighted_mktval, use = "complete.obs")
cor(CEO$log_weighted_profits, CEO$log_weighted_mktval, use = "complete.obs")
```

```{r}
hist(CEO$log_weighted_profits, main="Log of Weighted Profits", xlab = NULL)
hist(CEO$log_weighted_mktval, main="Log of Weighted Market Value", xlab = NULL)
```

<!-- Let's draw a scatterplot matrix for those weighted variables. -->

<!-- ```{r} -->
<!-- scatterplotMatrix( ~ log_salary + log_weighted_profits + log_weighted_mktval,  -->
<!--                    data = CEO,  -->
<!-- #                 reg.line="" ,  -->
<!--                    smoother="", -->
<!--                    diagonal = "histogram",  -->
<!--                    main = "Scatterplot Matrix for weighted profits / market value") -->
<!-- ``` -->

The correlation between `log_salary` and `log_weighted_profits` and `log_salary` and `log_weighted_mktval` aren't notably different from their unweighted counterparts. Thus, the number of years an individual has served as CEO doesn't greatly impact our analysis.

However, the correlation between `log_weighted_profits` and `log_weighted_mktval` is stronger than between the unweighted versions. This correlation could indicate that a longer tenured CEO's achieved profit has a larger effect on the company's market value.

## CEO Age

We perform a brief analysis of the impact of CEO age on our key variables. 

```{r}
hist(CEO$age, main = "CEO Age", xlab = NULL)
```

There is no significant skew in the age variable.

Next, we check whether there is a correlation between age and salary.

```{r}
cor(CEO$age, CEO$salary)
```

There is no notable linear relationship between age and salary. This can also be seen in the scatterplot.

```{r}
plot(CEO$age, CEO$salary)
```

Thus, we don't expect age to impact our primary analysis.

## CEO Education

Let's examine if there is a linear correlation between salary and education.

We add the education variables together to create one variable that indicates overall education and check the correlation.

```{r}
CEO$educ = CEO$college + CEO$grad
cor(CEO$educ, CEO$salary)
```

There is no notable linear correlation. Let's look at the boxplots for all education levels.

```{r}
educ_bin = cut(CEO$educ, breaks = 3, labels = 
      c("No College", "Attended College", "Attended Grad School"))
summary(educ_bin)
```

```{r}
boxplot(salary ~ educ_bin, data = CEO,
        main = "Salary by College Attendance",
        ylab = "salary in 1000 $s")
```

Since there are only 5 data points of CEOs with no college education, the "No College" boxplot has no significance and shall not be discussed further. The other two boxplots reveal that there is a very minor difference in the salary distribution between CEOs that attended college and those that attended grad school.

<!-- ## Seniority factor -->

<!-- ```{r} -->
<!-- hist(CEO$ceoten) -->
<!-- hist(CEO$comten) -->
<!-- hist(log(CEO$ceoten)) -->
<!-- ```  -->

<!-- The ceoten variable (amount of years the CEO has been in office within the company) is skewed. Use transformation????? -->

<!-- Finally, let's check out, if seniority and salary correlate in some way. -->

<!-- ```{r} -->
<!-- cor(CEO$comten, CEO$salary) -->
<!-- cor(CEO$ceoten, CEO$salary) -->

<!-- CEO$log_ceoten = log(CEO$ceoten) -->
<!-- CEO$log_ceoten[CEO$ceoten == 0] <- NA -->

<!-- cor(CEO$log_ceoten, CEO$log_salary, use = "complete.obs") -->

<!-- ``` -->

<!-- ```{r} -->
<!-- scatterplotMatrix( ~ salary + log_salary + ceoten + log_ceoten + comten,   -->
<!--                    data = CEO,  -->
<!-- #                 reg.line="" ,  -->
<!--                    smoother="", -->
<!--                    diagonal = "histogram",  -->
<!--                    main = "Scatterplot Matrix Seniority") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- scatterplotMatrix( ~ ceoten + comten + mktval + profits,   -->
<!--                    data = CEO,  -->
<!-- #                 reg.line="" ,  -->
<!--                    smoother="", -->
<!--                    diagonal = "histogram",  -->
<!--                    main = "Scatterplot Matrix Seniority") -->
<!-- ``` -->

## Confounding Variables

It is possible that there is a confounding variable effect in this dataset. If we look again at the scatterplot matrix for our primary variables, we can see that `salary` and `mktval` seem to be correlated. However, profits and mktval are also seem highly correlated.

```{r}
scatterplotMatrix(~salary + profits + mktval, data=CEO)
```

This relationship could imply that `mktval` is a confounding variable. It's reasonable that a company with a high market value posts larger profits than a company with a low market value. Similarly, a high market value company might decide to pay their CEO more money than a low market value company. In that way, market value could be confounding the relationship between profits and salary and producing a correlation that might not actually exist. This potential issue should be explored in a more rigorous analysis.

<!-- To get around this, we reduce the impact of mkval on profits by dividing profits by mktval. This should give us an intuitive variable that is the 1990 profits as a % of market value. As we can see from the histogram below, the variable is fairly normally distributed when profits are positive, with the negative profits observations being clear outliers. This can be seen more explicitly in the second histogram, where we are only looking at positive profit values. -->

<!-- ```{r} -->
<!-- CEO$prof_perc <- CEO$profits/CEO$mktval -->
<!-- hist(CEO$prof_perc,breaks=seq(-0.6,0.2,0.05),main="Histogram of Profits/Market-Value", xlab="Profits as a % of Market-Value") -->
<!-- hist(CEO$prof_perc[CEO$profits>0],main="Histogram of Profits/Market-Value (only positive values)", xlab="Profits as a % of Market-Value") -->
<!-- ``` -->

<!-- To ensure that the transformation has worked correctly, we can compare the new *prof_perc* variable to the *mktval* and *profits* variables.  -->

<!-- ```{r} -->
<!-- # scatterpot matrix of new prof_perc variable  -->
<!-- scatterplotMatrix(~prof_perc + profits + mktval ,data=CEO,   -->
<!--                   smoother="", -->
<!--                   diagonal="histogram", -->
<!--                   main="Scatterplot Matrix CEO Salary") -->

<!-- cor(CEO$prof_perc, CEO$mktval,use="complete.obs") -->
<!-- cor(CEO$prof_perc, CEO$profits,use="complete.obs") -->
<!-- # scatterplot matrix and correlations without negative values -->
<!-- scatterplotMatrix(~prof_perc + profits + mktval ,data=CEO[CEO$profits>0,],   -->
<!--                   smoother="", -->
<!--                   diagonal="histogram", -->
<!--                   main="Scatterplot Matrix CEO Salary") -->
<!-- cor(CEO$prof_perc[CEO$profits>0], CEO$mktval[CEO$profits>0],use="complete.obs") -->
<!-- cor(CEO$prof_perc[CEO$profits>0], CEO$profits[CEO$profits>0],use="complete.obs") -->
<!-- ``` -->

<!-- Finally, we can compare the new *prof_perc* variable to a few of the variables in our dataset. -->

<!-- ```{r} -->
<!-- # prof_perc vs other variables -->
<!-- scatterplotMatrix(~prof_perc + salary + ceoten + comten, data=CEO,  -->
<!--                   smoother="", -->
<!--                   diagonal="histogram", -->
<!--                   main="Scatterplot Matrix CEO Salary") -->

<!-- # prof_perc vs other variables with negative values removed -->
<!-- scatterplotMatrix(~prof_perc + ceoten + comten + salary ,data=CEO[CEO$profits>0,],  -->
<!--                   smoother="", -->
<!--                   diagonal="histogram", -->
<!--                   main="Scatterplot Matrix CEO Salary (Negative Profits Removed)") -->

<!-- # correlation with and without negative values -->
<!-- cor(CEO$prof_perc, CEO$salary, use="complete.obs") -->
<!-- cor(CEO$prof_perc[CEO$profits>0], CEO$salary[CEO$profits>0], use="complete.obs") -->
<!-- ``` -->

<!-- The new variable *prof_perc* has little to no correlation with the salary variable. This could indicate, that once the effects of market-value are removed there is very little correlation between profit and salary. This would need to be addressed further in a more rigorous approach. -->

# Discussion

In our analysis, we uncovered several features of this dataset that could be useful for statistical modeling. We noticed an overall positive relationship between a CEO's salary and the company's profits and market value. As we learn more exploratory techniques, we can further tune this analysis to better account for issues like confounding variables.